Modelling temporal dependence and multilevel structure using hidden Markov models

### 2020-08-06 Report/Sections/results.tex

We fit all four models from the simulation study to the data shown in Figure \ref{fig:data} to test their performance in a real-world setting. We first report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and then asses the quality of the fit. Finally, we compare the results of the CarHHMM-DFT with those of the other candidate models.

@@

The three subdive behaviours have distinct average ``wiggliness" values, as the means of $Z^{*(2)}$ for each state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean of $Z^{*(2)}$ and the smallest variance corresponding to $Z^{*(1)}$. It also has the highest auto-correlation in $Z^{*(1)}$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($Z^{*(2)}$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower in than subdive state 1. We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $Z^{*(2)}$ and variance of $Z^{*(1)}$ in subdive state 3 is much higher than the other two states. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower, implying more variation in acceleration between segments. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

@@

Two visual tools are used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}. To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ of a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $Z^{*(2)}$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.

We also plot histograms of $Y$ and $Z^*$ corresponding to each dive type or subdive state in Figure \ref{fig:empirical_dist}. Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}). Our results again mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. Another exception is $Z^{*(1)}$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}. See Section S-2.? for histograms for every observation and hidden state.

@@

The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3. In addition, the HHMM-DFT produces estimates of $\sigma_1^{*(1)}$, $\sigma_1^{*(2)}$, and $\sigma_1^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration. The estimated uncertainties of the estimates of $\mu_1^{*(1)}$, $\mu_1^{*(2)}$, and $\mu_1^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration. These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation. Further, the empirical auto-correlation of $Z^{*(1)}$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers. Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.

>>>>

We fit all four models from the simulation study to the data shown in Figure \ref{fig:data} to test their performance in a real-world setting. We first report the results from the full CarHHMM-DFT (defined in Section \ref{sec:data}) in detail and then asses the quality of the fit. We also compare the results of the CarHHMM-DFT with those of the other candidate models.

@@

The three subdive behaviours have distinct average ``wiggliness" values, as the means of $\Ztwo$ for each state are separated by an order of magnitude (Table \ref{table:emis_dists_CarHHMM-DFT}, Figure \ref{fig:fine_emis}). 
Subdive state 1 has the smallest mean of $\Ztwo$ and the smallest variance corresponding to $\Zone$. It also has the highest auto-correlation in $\Zone$. This implies less overall activity and more consistent acceleration compared to the other subdive states. 
Subdive state 2 has a mean ``wiggliness" ($\Ztwo$) that is one order of magnitude higher than subdive state 1 and its acceleration has about twice the variance compared to subdive state 1. The auto-correlation of acceleration is also slightly lower in than subdive state 1. We therefore hypothesize that subdive type 2 corresponds to fluking (active swimming), as strong sinusoidal behaviour in acceleration is characteristic of this behaviour in marine mammals \citep{Simon:2012}.
Finally, the mean of $\Ztwo$ and variance of $\Zone$ in subdive state 3 is much higher than the other two states. The auto-correlation of $\Zone$ is also much lower, implying more variation in acceleration between segments. This corresponds to vigorous swimming activity, especially as the killer whale begins or ends a dive (see Figure \ref{fig:labeled_dives}).

@@

Two visual tools are used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual is the marginal cumulative distribution function of an observation conditioned on all other observations \citep{Zucchini:2016}. To easily visualize outliers, pseudo-residuals are often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ of a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. Histograms of the pseudoresiduals mostly support that this model is well-specified. One exception is $\Ztwo$, whose pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $\Ztwo$ may follow a heavier-tailed distribution than the gamma distribution used in the case study.

We also plot histograms of $Y$ and $Z^*$ corresponding to each dive type or subdive state in Figure \ref{fig:empirical_dist}. Observations are weighted by the decoded probability that they correspond to a particular hidden state, and this histogram is plotted over the emission distribution learned by the CarHHMM-DFT (see Figure \ref{fig:empirical_dist}). Our results again mostly support a well-specified model with the exception of $\Ztwo$, which is again right-skewed. Another exception is $\Zone$, which has heavy tails in subdive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study and may indicate biologically relevant phenomena such as prey capture \citep{Tennessen:2019a}. See Section S-2.? for histograms for every observation and hidden state.

@@

The HHMM-DFT decodes dive types and subdive states similarly to the CarHHMM-DFT, but it is less likely to categorize the behaviour at the beginning and end of dives as subdive type 3. In addition, the HHMM-DFT produces estimates of $\sigma_A^{*(1)}$, $\sigma_A^{*(2)}$, and $\sigma_A^{*(3)}$ (the standard deviation of acceleration) which are approximately 50-100 \% larger than those of the CarHHMM-DFT across all axes of acceleration. The estimated uncertainties of the estimates of $\mu_A^{*(1)}$, $\mu_A^{*(2)}$, and $\mu_A^{*(3)}$ are also less than half of those for the CarHHMM-DFT across all axes of acceleration. These findings are consistent with the tendency to overestimate variance and underestimate standard error when ignoring auto-correlation. Further, the empirical auto-correlation of $\Zone$ within each decoded subdive state was at least 0.5 for all acceleration axes and subdive states, and the pseudoresiduals of the HHMM-DFT contain several outliers. Both of these findings suggest that the HHMM-DFT is a significantly worse fit to this data than the full CarHHMM-DFT.

### 2020-08-06 Report/main.tex
Updates from Overleaf
Update Report
Stylistic changes to report

% Title, authors, affiliations
\title[]{Modelling fine-scale correlation structures using hidden Markov models}%\query{Q1}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

@@

Recent advances in high-frequency tagging
technology have made a vast
amount of movement data available in a variety of fields. This data can exhibit simultaneous behavioural processes occurring at different
time scales, resulting in increasingly complicated dependence structures. These processes can be modelled 
through a hierarchical hidden Markov model
(HHMM), which models the system as a
nested structure of hidden Markov models (HMMs).
At very short time scales, however, many basic assumptions of traditional HMMs are violated. We demonstrate how
to incorporate fine-scale processes into the
larger structure of HHMMs while maintaining computational efficiency. We apply our
method to dive and accelerometer data collected from a northern resident killer whale off
the coast of British Columbia, Canada.

@@

%%%%%%%%%%%%
%\vfill\eject
%{\centerline{\Large{Notes from Nancy}}}
%\input{Sections/0627NH}
%\vfill\eject
%%%%%%%%%%%%%
%
\section{Introduction}
\input Sections/intro.tex
%
\section{Models and parameter estimation}
\input Sections/models.tex
%
\section{Simulation Study}
\input Sections/simulation.tex
%
\section{Killer Whale Case Study}
\label{sec:case_study}
\input Sections/results.tex
%
\section{Discussion}
\input Sections/discussion.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{Figures and Tables}
\input Sections/figures.tex
%
%\CJShistory

>>>>

% Title, authors, affiliations
\title[]{Modelling temporal dependence and multilevel structure using hidden Markov models}%\query{Q1}
\author{BlindedA\authorref{1}\thanksref{*}}
\author{BlindedB\authorref{2}}
\affiliation[1]{Author affiliations will go here in the accepted manuscript, 
but do NOT include them in your initial submission because it must be anonymous.}
\affiliation[2]{Second Affiliation}

@@

Advances in tagging technology have made a vast amount of high-frequency time series data available in several fields. These data often exhibit complicated temporal dependence structure which cannot be modelled using common methods of Functional Data Analysis (FDA). The field of animal movement suggests a hierarchical approach using hidden Markov models (HMMs), but many intricate fine-scale processes violate basic assumptions of HMMs.
We detail a generalization of the hierarchical HMM approach that incorporates a wider class of parametric models at problematic fine scales. 
%We also provide several examples of viable fine-scale models.
As a case study, we build a hierarchical model that combines an HMM with a moving-window auto-regressive process to describe the movement of a killer whale off the western coast of Canada. 
Our model outputs more interpretable results as well as more accurate parameter and standard error estimates compared to existing models.

@@

%%%%%%%%%%%%
%\vfill\eject
%{\centerline{\Large{Notes from Nancy}}}
%\input{Sections/0627NH}
%\vfill\eject
%%%%%%%%%%%%%
%
\section{Introduction}
\input Sections/intro.tex
%
\section{Models and parameter estimation}
\label{sec:models}
\input Sections/models.tex
%
\section{Data Collection and Model Construction}
\label{sec:data}
\input Sections/data.tex
%
\section{Simulation Study}
\label{sec:sim_study}
\input Sections/simulation.tex
%
\section{Case Study Results}
\label{sec:case_study}
\input Sections/results.tex
%%
\section{Discussion}
\input Sections/discussion.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{Figures and Tables}
\input Sections/figures.tex
%
%\CJShistory
